---
title: "RepData_PeerAssessment2.Rmd"
author: "Jennifer Campbell"
date: "May 29, 2017"
output: 
  html_document: 
    keep_md: yes
---

# Impact of Severe Weather Events on the United States

## Synopsis
This analysis investigates the impact of severe weather events on the United States, specifically their effect on public health and their economic consequences. Raw data is provided by the U.S. National Oceanic and Atmospheric Administration's (NOAA). The findings below show that ___



## Data Processing
In the first step of the analysis, check whether the input data file activity.csv is in the working directory, and then download and unzip the file if necessary:

``` {r, echo=TRUE}
if(!file.exists("storm.csv.bz2")){
  url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
  download.file(url, destfile='storm.csv.bz2', method="wininet", mode="wb")
}
```

Then we can read in the data:  
```{r echo=TRUE}
df <- read.csv("storm.csv.bz2", 
               na.strings = "NA", header = TRUE)
dim(df)
names(df)
```                

There are a total of 902297 observations of 37 variables in this dataset.  

### Wrangling the Data
In order to work more efficiently, we will subset the data to only include those years and variables which we are most interested in.  
The data include events from 1950 to 2011, but in the earlier years, there are generally fewer events recorded due to a lack of good records, as shown in the histogram of Total Storm Events by Year:

```{r echo=TRUE}
df$Year <- as.numeric(format(as.Date(df$BGN_DATE, format = "%m/%d/%Y %H:%M:%S"), "%Y"))
hist(df$Year, main="Total Storm Events by Year", xlab="Year")
```

We can see a marked increase in the number of recorded storm events after the year 1990. Therefore, to take advantage of the best records, we will focus our examination on the subset of strom data between 1990 and 2011.

In examining our questions, the variables which we will be using in this dataset are:

### Codebook  
| Field | Description|
|:--------|:---------------------|
|EVTYPE | The type of weather event (i.e. flood, tornado; a full listing can be found on page 6 of the  [Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf) under section 2.1.1 Storm Data Event Table) <br> Factor  variable; No missing ("NA") values|
|FATALITIES | The number of fatalities caused by the weather event <br> Numeric variable, Range: 0-583; No missing values|
|INJURIES | The number of injuries caused by the weather event <br> Numeric variable, Range: 0-1568; No missing values|
|PROPDMG | The estimated property damage caused by the weather event in Dollars <br> Numeric variable, No missing values|
|PROPDMGEXP | The order of magnitude of property damage <br> Factor variable, No missing values|
|CROPDMG | The estimated crop damage caused by the weather event in Dollars <br> Numeric variable, No missing values|
|CROPDMGEXP | The order of magnitude of crop damage <br> Factor variable <br> No missing values|

Thus, the new subsetted dataframe can be defined:

```{r echo=TRUE}
stormdata <- df[df$Year %in% c(1990:2011), ]
variables <- c("EVTYPE", "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP", 
                "CROPDMG", "CROPDMGEXP")
stormdata <- stormdata[variables]
```

so that we are now working with 751,740 observations of 7 variables.

The final step in preparing the data will be to clean up some of the text variables. As visible when using `summary()` on the EVTYPE factor variable, many of the names have been entered with character variations such as lowercase and plural usage, or the abbreviation "TSTM" for thunderstorm. Making these entries more consistant will provide a better overall picture of the health and economic data as functions of the type of weather event.

```{r echo=TRUE}
stormdata$EVTYPE <- toupper(stormdata$EVTYPE)
stormdata$EVTYPE <- sub("TSTM", "THUNDERSTORM", stormdata$EVTYPE)
stormdata$EVTYPE <- sub("S$", "", stormdata$EVTYPE)
stormdata$EVTYPE <- sub("WINDS", "WIND", stormdata$EVTYPE)
stormdata$EVTYPE <- sub("FLOODING", "FLOOD", stormdata$EVTYPE)
stormdata$EVTYPE <- sub("FLD", "FLOOD", stormdata$EVTYPE)
stormdata$EVTYPE <- sub("SML", "SMALL", stormdata$EVTYPE)
stormdata$EVTYPE <- sub("WIND/HAIL", "WIND HAIL", stormdata$EVTYPE)
stormdata$EVTYPE <- as.factor(stormdata$EVTYPE)
```

## Question: Across the United States, which types of events are most harmful with respect to population health?
  
The impact of storm events on population health can be described in terms of the numbers of fatalaties and injuries caused. Here, we will look at the sum of all fatalaties and the sum of all injuries for each weather event type, and rank the top 10:

```{r echo=TRUE}
library(stats)
library(plyr)
sumFatalities <- aggregate(stormdata$FATALITIES, 
                           by = list(stormdata$EVTYPE), FUN = "sum")
sumFatalities <- arrange(sumFatalities, sumFatalities[, 2], decreasing = T)
sumInjuries <- aggregate(stormdata$INJURIES, 
                           by = list(stormdata$EVTYPE), FUN = "sum")
sumInjuries <- arrange(sumInjuries, sumInjuries[, 2], decreasing = T)

topFatalities <- head(sumFatalities, n = 10)
topInjuries <- head(sumInjuries, n = 10)
names(topFatalities) <- c("EventType", "Fatalities")
names(topInjuries) <- c("EventType", "Injuries")

topFatalities
topInjuries
```
 
These tables show us that the weather events with the highest impact (as determined by the highest total numbers of deaths and injuries caused) include Excessive Heat, Tornados, Thunderstorm Winds, Flash Floods and Lightning, and that these two lists of events hold most of their members in common.  
 
## Question: Across the United States, which types of events have the greatest economic consequences?
 
The economic consequences of weather events can be quantified by the estimated property and crop damage.   

```{r echo=TRUE}

```

 
## Results  

The following panel plot contains a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).

```{r echo=TRUE}
par(mfrow = c(1,2), mar = c(10, 4, 2, 1), cex = 0.8)
barplot(topFatalities$Fatalities, names = topFatalities$EventType, las = 3,
                     ylab = "Total Fatalities", main = "10 Most Fatal Weather Events")
barplot(topInjuries$Injuries, names = topInjuries$EventType,las = 3, 
                     ylab = "Total Injuries", main = "10 Most Injurious Weather Events")

```

The plot shows that the pattern of activity follows the same trend across weekdays and weekends, with a peak in the morning (as we saw previously around 8:35AM), but that weekends tend to have higher average step counts throughout the day than on weekdays. 
The output of `summary()` for each day type also confirms this trend of higher weekend averages compared to weekdays, though the weekdays have a higher maximum.  

```{r echo=TRUE}


```

The weekday data appears closer to the original estimates, which makes sense, as it makes up a greater portion of the total dataset.
(Also of note: there are fewer weekend observations in the data set, and therefore fewer values to average across.)

*End of analysis*
